An exposition on the propriety of restricted Boltzmann machines

### 2016-04-20 writing/degeneracy/draft.Rmd
restructured the degeneracy section

---
title: "Degeneracy in a restricted Boltzmann machine"
author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
date: "January 8, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: ../references/refs.bib
---

@@

#Degeneracy in Random Graph Models
In Random Graph Model theory, \emph{degeneracy} occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. We consider three related definitions of degeneracy as they relate to a graphical model. Section \ref{mean-parameterization} presents a definition from @handcock2003assessing based on the mean parameterization of the model parameters, section \ref{neg-potential} presents a definition from @schweinberger2011instability based on the neg-potential function, and section \ref{conditional-expectation} presents a definition from @kaiser2006markov that involves the difference between expectations of the model and expectations under independence.

##Mean Parameterization
@handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}.

@@

##Neg-potential
@schweinberger2011instability introduces a related concept, instability, which  is characterized
by excessive sensitivity and near-degeneracy. He provides the following tool for detecting instability in the a discrete exponential family model, and thus near-degeneracy.

@@

\begin{lemma}
If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X_N \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N}$ is the subset of $\epsilon$-modes of the probability mass function.
\end{lemma}

In the context of the RBM, $N$ is the number of nodes in the model, $H + V$. So, the distribution is unstable if for any $C > 0$,

\begin{align*}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] &= \max\limits_{\boldsymbol x \in \mathcal{X}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
&= \max\limits_{h_j, v_i \in \{0,1\}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
&> C(H + V)
\end{align*}

There are $H + V + H*V$ terms in the sum that comprises $Q(\boldsymbol x)$, which leads to a natural split of the quantity into main effects, those terms based solely on $h_i$ and $v_j$ and interaction effects, those terms based on the product of $h_i v_j$. There are $H+V$ main effects terms and $H*V$ interaction effects terms. For $H$ and $V$ large enough, the number of interaction terms is growing much faster than $H + V$, meaning no matter the $C$, we are able to find values of $\boldsymbol theta$ that can make the left side of equation \ref{eq:unstable} large compared to $C(H+V)$. This points to the interaction terms as potentially the more dangerous parameters for causing near-degeneracy in the RBM.

@@

##Expectation Differences
The characterization put forth by @kaiser2006markov of degeneracy in a model concerns the difference between model expectations and expectations given independence. Using this concept of dependence in a model, it is possible to understand what conditions lead to near-degeneracy in a model versus those that guarantee non-degenerate models. In this method, a model is considered without near-degeneracy if its expected value is not very different from the same model with assumed independence. The quantities to compare in the RBM case are

\begin{align*}
\text{E}\left[\boldsymbol X\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}

#Tiny example
To illustrate the idea of graph degeneracy in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

@@

##Degeneracy three ways

>>>>

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
date: "May, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    number_sections: yes
bibliography: ../references/refs.bib
---

@@

#Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of the RBM ($H + V + H V$ parameters) makes three characteristics of model impropriety of particular concern, *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, discuss how to detect them in the RBM, and point out their relationship to one another.

##Near-degeneracy
In Random Graph Model theory, *degeneracy* occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. @handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}. In other words, if random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$ then the model is near-degenerate, as defined by @handcock2003assessing.

@@

##Instability
@schweinberger2011instability introduces a related concept, *instability*, which occurs when small changes in the natural parameters result in large changes in probability masses. This can be thought of as excessive sensitivity in the model. @schweinberger2011instability provides the following tool for detecting instability in the a discrete exponential family model.

@@

Conversely, a discrete exponential family distribution is stable if there exists constants $C > 0$ and $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] \le CN \text{ for all } N > N_C. \label{eq:stable}
\end{align}

Within the context of the RBM, $N$ is the number of nodes in the model, $H + V$. To ensure the model is stable, we must restrict $\boldsymbol \theta$ such that there exists constants $C > 0$ and $N_C > 0$ that force equation \ref{eq:stable} to hold.

\begin{align*}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] &= \max\limits_{\boldsymbol x \in \mathcal{X}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
&= \max\limits_{h_j, v_i \in \{0,1\}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
&\le C(H + V)
\end{align*}

There are $H + V + H*V$ terms in the sum that comprises $Q(\boldsymbol x)$, which leads to a natural split of the quantity into main effects, those terms based solely on $h_i$ and $v_j$ and interaction effects, those terms based on the product of $h_i v_j$. There are $H+V$ main effects terms and $H*V$ interaction effects terms. For $H$ and $V$ large enough, the number of interaction terms is growing much faster than $H + V$, meaning no matter the $C$, we can restrict the size of the left side of equation \ref{eq:stable} effectively by restricting the size of the interaction terms. This points to the interaction terms as potentially the more dangerous parameters for causing instability in the RBM simply because there are so many more of them.

@@

Instability in a model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy.

\begin{lemma}[Schweinberger (2011)]
If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X_N \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N}$ is the subset of $\epsilon$-modes of the probability mass function.
\end{lemma}

##Uninterpretability
@kaiser2007statistical defines a measure of model impropriety called *uninterpretability*, which is characterized as due to the existence of dependence, marginal mean-structure no longer being maintained in the model. This measure concerns the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$. Using this concept of dependence in a model, it is possible to understand what conditions lead to uninterpretability in a model versus those that guarantee interpretable models. In this method, a model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence. The quantities to compare in the RBM case are

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}
and
\begin{align*}
\text{E}\left[\boldsymbol X | \emptyset \right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}

#Tiny example
To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

@@

##Impropriety three ways
